Recurrent Deep Learning Models and its applications

### 2025-02-09 paper2/main.tex
paper update

\newpage
\subsubsection{Hidden Markov Model}
Before reviewing Hidden Markov Model (HMM), it is essential to understand what Markov models is, or Markov chains. Markov chains are fundamental models in probability theory and statistic, and it is a stochastic process describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. This property known as the Markov property. In rigorous terms, let the state space be defined as:
Let the state space be defined as:
\begin{equation}
    S = \{ s_1, s_2, \dots, s_N \}.
\end{equation}
\begin{equation}
    P(s_{t+1} \mid s_t, s_{t-1}, s_{t-2}, \cdots, s_{1}) = P(s_{t+1} \mid s_t)
\end{equation}
The state transition probability is denoted by:
\begin{equation}
    P_{ij} = P(s_{t+1} = s_j \mid s_t = s_i).
\end{equation}

Given an initial state \( s_1 \) with probability \( P(s_1) \), the joint probability of a state sequence \( \{s_1, s_2, \dots, s_T\} \) can be written as:
\begin{equation}
    P(s_1, s_2, \dots, s_T) = P(s_1) \cdot P(s_2 \mid s_1) \cdot P(s_3 \mid s_2) \cdots P(s_T \mid s_{T-1})
\end{equation}
We can simplfy the joint probability of a state sequence as equestion 47.
\begin{equation}
    P(s_1, s_2, \dots, s_T) = P(s_1) \prod_{t=1}^{T-1} P(s_{t+1} \mid s_{t})
\end{equation}

In the context of Natural Language Processing (NLP), early language models utilized the Markov assumption by approximating the probability of a word sequence. For example, for a bigram model, the probability of a sentence \( w_1, w_2, \dots, w_T \) is given by:
$$
P(w_1, w_2, \dots, w_T) \approx \prod_{t=1}^{T} P(w_t \mid w_{t-1}),
$$
where each word is assumed to depend only on its immediate predecessor. Although this simplification makes computation manageable, it does not account for deeper syntactic or semantic structures inherent in language.

Literature Review on Markov Chains

Markov chains are fundamental models in probability theory that describe systems evolving in discrete time where the future behaviour of the process depends solely on its current state—a property known as the **Markov property**. In rigorous terms, let \((\mathcal{X}, \mathcal{F})\) be a measurable space and let \(\{X_n\}_{n\ge0}\) be a sequence of random variables taking values in \(\mathcal{X}\). The process is said to be a *Markov chain* if, for every \(n\ge0\) and for every \(A\in\mathcal{F}\),
\[
P\bigl(X_{n+1}\in A \mid X_n, X_{n-1}, \dots, X_0\bigr) = P\bigl(X_{n+1}\in A \mid X_n\bigr).
\]
This memoryless property implies that the current state \(X_n\) fully encapsulates all the probabilistic information necessary for predicting the future evolution of the process.

Stationarity and Convergence

A probability distribution \(\pi\) on \(\mathcal{X}\) is called *stationary* (or invariant) for the Markov chain if it satisfies
\[
\pi(A) = \int_{\mathcal{X}}P(x,A)\,\pi(dx) \quad \text{for all } A\in\mathcal{F},
\]
where \(P(x,A)\) denotes the one-step transition probability. Under appropriate conditions—most notably, **\(\varphi\)-irreducibility** and **aperiodicity**—it can be rigorously shown that the distribution of the chain converges to \(\pi\) in total variation distance. Specifically, for \(\pi\)-almost every initial state \(x\),
\[
\lim_{n\to\infty} \|P^n(x,\cdot)-\pi(\cdot)\|_{\text{TV}} = 0,
\]
where the total variation distance between two probability measures \(\mu\) and \(\nu\) is defined as
\[
\|\mu-\nu\|_{\text{TV}} = \sup_{A\in\mathcal{F}}|\mu(A)-\nu(A)|.
\]
Such convergence results are established in depth in Roberts and Rosenthal (2004) and form the theoretical backbone for many modern computational methods.

Small Sets, Drift Conditions, and Geometric Ergodicity

For Markov chains on general (and possibly uncountable) state spaces, it is crucial to establish conditions that guarantee not only convergence but also quantitative rates of convergence. One such condition involves the concept of a **small set**. A subset \(C\subseteq \mathcal{X}\) is called small if there exist a positive integer \(n_0\), a constant \(\epsilon>0\), and a probability measure \(\nu\) on \(\mathcal{X}\) such that
\[
P^{n_0}(x,A) \ge \epsilon\,\nu(A), \quad \text{for all } x\in C \text{ and all } A\in\mathcal{F}.
\]
This minorisation condition is instrumental in establishing uniform convergence bounds.

Complementary to the small set condition is a **drift condition**. A drift (or Lyapunov) function \(V:\mathcal{X}\to[1,\infty)\) is used to quantify the tendency of the chain to return to a particular subset \(C\). The chain is said to satisfy a drift condition if there exist constants \(\lambda < 1\) and \(b < \infty\) such that
\[
P V(x) \le \lambda\, V(x) + b\,1_C(x) \quad \text{for all } x\in\mathcal{X},
\]
where \(P V(x) = \int_{\mathcal{X}} V(y)P(x,dy)\) and \(1_C\) is the indicator function of \(C\). When a Markov chain satisfies both a minorisation condition and an appropriate drift condition, it is often **geometrically ergodic**; that is, there exist a function \(M(x)\) and a constant \(\rho\in(0,1)\) such that
\[
\|P^n(x,\cdot)-\pi(\cdot)\|_{\text{TV}} \le M(x)\,\rho^n, \quad \text{for all } n\ge0.
\]
Such quantitative convergence rates are crucial for both theoretical investigations and practical applications (see Roberts and Rosenthal, 2004; Meyn and Tweedie, 1993).

Coupling, Regeneration, and Quantitative Bounds

A powerful method for analyzing the convergence of Markov chains is the use of **coupling techniques**. In this approach, one constructs a joint process \((X_n, Y_n)\) on \(\mathcal{X}\times\mathcal{X}\) such that:
- The marginal dynamics of \(X_n\) and \(Y_n\) each follow the prescribed transition kernel \(P\).
- The coupling is designed so that the probability \(P(X_n\neq Y_n)\) decreases over time.

A key consequence of coupling is the following inequality:
\[
\|P^n(x,\cdot)-\pi(\cdot)\|_{\text{TV}} \le P(X_n \neq Y_n).
\]
By constructing couplings that "meet" (or "couple") quickly, one may deduce explicit bounds on the rate of convergence in total variation. Similar ideas underpin **regeneration techniques**, which partition the Markov chain into independent, identically distributed tours. Regeneration points allow the application of classical renewal theory to derive central limit theorems (CLTs) for additive functionals of the chain. For example, if \(h:\mathcal{X}\to\mathbb{R}\) is a function with \(\pi(|h|^2)<\infty\), then under appropriate conditions
\[
\frac{1}{\sqrt{n}}\sum_{i=0}^{n-1}\!\Bigl(h(X_i)-\pi(h)\Bigr) \xrightarrow{d} \mathcal{N}(0,\sigma^2),
\]
with the asymptotic variance \(\sigma^2\) often expressible in terms of the chain’s autocovariance structure.

>>>>

\newpage
\subsubsection{Hidden Markov Model}
Before reviewing Hidden Markov Model (HMM), it is essential to understand what Markov models is, or Markov chains. Markov chains are fundamental models in probability theory and statistic, and it is a stochastic process describing a sequence of possible events in which the probability of each event depends only on the state attained in the previous event. This property known as the Markov property \parencite{roberts2004general}. In rigorous terms, let the state space be defined as:
Let the state space be defined as:
\begin{equation}
    S = \{ s_1, s_2, \dots, s_N \}.
\end{equation}
\begin{equation}
    P(s_{t+1} \mid s_t, s_{t-1}, s_{t-2}, \cdots, s_{1}) = P(s_{t+1} \mid s_t)
\end{equation}
The state transition probability is denoted by:
\begin{equation}
    P_{ij} = P(s_{t+1} = s_j \mid s_t = s_i).
\end{equation}

Given an initial state \( s_1 \) with probability \( P(s_1) \), the joint probability of a state sequence \( \{s_1, s_2, \dots, s_T\} \) can be written as:
\begin{equation}
    P(s_1, s_2, \dots, s_T) = P(s_1) \cdot P(s_2 \mid s_1) \cdot P(s_3 \mid s_2) \cdots P(s_T \mid s_{T-1})
\end{equation}
We can simplfy the joint probability of a state sequence as equestion 47.
\begin{equation}
    P(s_1, s_2, \dots, s_T) = P(s_1) \prod_{t=1}^{T-1} P(s_{t+1} \mid s_{t})
\end{equation}
Then it is essential to discuss marginal probability in a Markov model because it provides critical information about the likelihood of being in particular state $s_j$ at a specific time $t$. The marginal probability of being in state $s_j$ at time $t$ defined as:
\begin{equation}
    \pi_t(s_j) = P(s_t=s_j)
\end{equation}
It is the probability of the system being in state $s_j$ at time t, irrespective of how the system transitioned to $s_j$. We can obtainde by summing over all possible transition probability $P_{ij}$ where $s_i$ belong in state space. 
\begin{equation}
    \pi_{t+1}(s_j) = \sum_{s_i \in S} \pi_t(s_i) \cdot P_{ij}
\end{equation}
If Markov chain is irreducible and aperiodic, the stationary distribution defined as below:
\begin{equation}
    \pi_j = \sum_{i=1}^{n} \pi_i P_{ij}
\end{equation}

\newpage
\textbf{Hidden markov model}
\\[1ex]
Hidden Markov Model (HMMs) is a statistical model that extends the Markov chains by introducting a layer of latent (unobservable) variables. And connecting them to observable outputs (emission probability). In contrast to standard Markov chains, hidden states are never be observed directly. Instead, we observe a sequence of data (observations) that are probabilitically generated by these hidden state. This structure has proven useful for applications in speech recognition, biological sequence analysis, and signal processing \parencite{rabiner1989atutorial}. Lets defined the 
% ####################################################
\newpage

Literature Review on Hidden Markov Models

Hidden Markov Models (HMMs) extend classical Markov chains by incorporating an additional layer of latent (unobservable) variables that govern the generation of the observed data. In contrast to standard Markov chains—where the state of the system is directly observed—an HMM assumes that the underlying process is hidden and that each observed datum is stochastically generated from the corresponding hidden state. This two-layer structure has proven to be both theoretically interesting and practically useful for applications in speech recognition, biological sequence analysis, and signal processing (Rabiner, 1989).

1. From Markov Chains to Hidden Markov Models

Let \(\{X_n\}_{n \ge 1}\) be a Markov chain on a measurable space \((\mathcal{X}, \mathcal{F})\) satisfying
\[
P\bigl(X_{n+1} \in A \mid X_n, X_{n-1}, \dots, X_1\bigr) = P\bigl(X_{n+1} \in A \mid X_n\bigr) \quad \text{for all } A \in \mathcal{F}.
\]
This *Markov property* ensures that the future evolution of the process depends only on its present state, not on the full past history.

In extending this framework to HMMs, we introduce an observable process \(\{Y_n\}_{n \ge 1}\) defined on another measurable space \((\mathcal{Y}, \mathcal{G})\). The connection between the hidden states and the observations is governed by the following assumption:

- **Conditional Independence of Observations:** Given the current hidden state,
  \[
  P\bigl(Y_n \in B \mid X_n, X_{n-1}, \dots, X_1, Y_{n-1}, \dots, Y_1\bigr) = P\bigl(Y_n \in B \mid X_n\bigr)
  \]
  for every \(B \in \mathcal{G}\). This states that the observation \(Y_n\) is independent of past states and observations, conditioned on \(X_n\).

Thus, an HMM is fully specified by the following three components:
1. The **initial distribution** \(\pi\) on \(\mathcal{X}\):
   \[
   \pi(x) = P(X_1 = x).
   \]
2. The **transition kernel** \(P(x, dy)\) of the hidden Markov chain:
   \[
   p_{ij} = P(X_{n+1} = j \mid X_n = i).
   \]
3. The **emission (observation) distribution** \(B(x, dy)\) or \(b_x(y)\):
   \[
   b_x(y) = P(Y_n = y \mid X_n = x).
   \]

2. Joint and Marginal Distributions

Using the above structure, the joint probability of a sequence of hidden states and observations \(\{(X_1, Y_1), \dots, (X_N, Y_N)\}\) is given by
\[
\begin{aligned}
P\bigl(X_1, \dots, X_N,\, Y_1, \dots, Y_N\bigr) 
&= \pi(X_1) \, b_{X_1}(Y_1) \prod_{n=2}^N P\bigl(X_n \mid X_{n-1}\bigr) \, b_{X_n}(Y_n)\\[1mm]
&= \pi(X_1)\, b_{X_1}(Y_1)\, p_{X_1X_2}\, b_{X_2}(Y_2) \cdots p_{X_{N-1}X_N}\, b_{X_N}(Y_N).
\end{aligned}
\]
Marginalization over the hidden states yields the likelihood of the observations:
\[
P(Y_1, \dots, Y_N) = \sum_{X_1 \in \mathcal{X}} \cdots \sum_{X_N \in \mathcal{X}} \pi(X_1)\, \prod_{n=2}^N p_{X_{n-1}X_n}\, \prod_{n=1}^N b_{X_n}(Y_n),
\]
which in many applications is used for parameter estimation and inference.

3. Inference and Algorithmic Developments

Due to the latent structure in HMMs, direct computation of the likelihood or the posterior distribution of the hidden states is usually intractable when the state space is large or continuous. This has spurred the development of several efficient algorithms:

- **The Forward-Backward Algorithm:** Utilizes dynamic programming to recursively compute the probabilities of partial observation sequences. For instance, denoting the forward variable by
  \[
  \alpha_n(x) = P(Y_1, Y_2, \dots, Y_n, X_n = x),
  \]
  one can compute
  \[
  \alpha_{n+1}(y) = \left[\sum_{x \in \mathcal{X}} \alpha_n(x) \, p_{xy}\right]\, b_y(Y_{n+1}).
  \]

- **The Viterbi Algorithm:** Determines the most probable sequence of hidden states (i.e., finds \(\hat{X}_1, \dots, \hat{X}_N\) such that the joint probability is maximized).

- **The Baum-Welch Algorithm (Expectation-Maximization):** Iteratively estimates the unknown parameters (transition probabilities and emission probabilities) of the HMM when they are not known a priori.

These methods rely on the underlying assumption that the hidden state process is a Markov chain, and many of the rigorous convergence and mixing properties discussed in the literature on general state space Markov chains (e.g., Roberts and Rosenthal, 2004; Meyn and Tweedie, 1993) carry over—often with additional considerations due to the unobserved structure.

4. Theoretical Underpinnings and Convergence

As with observable Markov chains, the study of HMMs involves rigorous mathematical analysis centered on convergence properties and ergodicity. Under appropriate conditions (e.g., the hidden chain is \(\varphi\)-irreducible, aperiodic, and satisfies a suitable drift condition), one can prove that the filtering distributions
\[
P(X_n \in \cdot \mid Y_1, \dots, Y_n)
\]
converge to a unique invariant measure. Techniques such as **coupling** and **regeneration**—which have been successfully employed in establishing quantitative convergence rates for Markov chains (Roberts \& Rosenthal, 2004)—are also adapted to the HMM context to study the stability and consistency of inference procedures.

For example, under the standard assumptions of an HMM, if the hidden state chain is geometrically ergodic then for a function \(h : \mathcal{X} \to \mathbb{R}\) with finite second moment under the stationary measure \(\pi\), the central limit theorem holds for the additive functional
\[
\frac{1}{\sqrt{n}} \sum_{i=1}^n \left[h(X_i) - \pi(h)\right],
\]
which is essential for quantifying the uncertainty in estimators derived via the Expectation-Maximization or sequential Monte Carlo methods (see also Billingsley, 1995).

5. Conclusion

In summary, by extending the framework of Markov chains—characterized by the Markov property, established convergence criteria, small set theory, and drift conditions—to a hidden state setting, Hidden Markov Models provide a robust foundation for dealing with complex sequence data where the underlying process is not directly observable. This extension leads to rich algorithmic and inferential techniques that have been rigorously analyzed and widely adopted in several applied fields.
% ######################## Background and Literature Review #########################
